---
title: "UNESCO_WDPA"
author: "Nick McManus"
date: "7/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)  # polygons
library(raster) #raster package
library(terra)  # newer/faster/better raster package
library(rnaturalearth)  #administrative boundaries data
library(exactextractr)  # extracting what's underneath polys, points
library(kableExtra)
library(tidyverse)
```

## Read in Data

As the country with the second highest number of UNESCO world heritage sites, we'll examine China

```{r}
# get china from rnaturalearth data
china <- ne_countries(
  scale = "medium",
  country = "China",
  returnclass = "sf") %>% 
  dplyr::select(sovereignt)

# plot
ggplot() +
  geom_sf(data = china)
```

```{r}
# define crs so it's the same across everything
my_crs <- st_crs(china)
```


## WDPA protected areas
UNEP-WCMC and IUCN (2021), Protected Planet: The World Database on Protected Areas (WDPA) [On-line], Cambridge, UK: UNEP-WCMC and IUCN. Available at: www.protectedplanet.net
```{r}
# read in raster with terra::rast
wdpa_global <- rast("data/wdpa_raster.tif") 

# transform china to crop bc it's faster than re projecting the whole global raster
china_trans <- china %>% 
  st_transform(crs = crs(wdpa_global)) %>% 
  vect() # vect is more agreeable to rasts

# crop and mask chen to china then project back to our crs
wdpa <- wdpa_global %>% 
  crop(china_trans) %>% 
  mask(china_trans) %>% 
  terra::project(y = crs(china))

plot(wdpa)
```
## Urban areas
Current (2015) areas classified as urban under Chen et al., 2022
Chen, G., Li, X., & Liu, X. (2022). Global land projection based on plant functional types with a 1-km resolution under socio-climatic scenarios. Scientific Data, 9(1), 125. https://doi.org/10.1038/s41597-022-01208-6
```{r}
chen <- rast("data/global_PFT_2015.tif") 

# transform italy to crop bc it's faster than re projecting the whole global raster
china_trans <- china %>% 
  st_transform(crs = crs(chen)) %>% 
  vect() # vect is more agreeable to rasts

# crop and mask chen to italy then project back to our crs
chen_cropped <- chen %>% 
  crop(china_trans) %>% 
  mask(china_trans) %>% 
  terra::project(y = crs(china))

# isolate just urban
urban <- chen_cropped
urban[urban != 19] <- 0 # if value isn't 19, make 0
urban[urban == 19] <- 1 # if value is 19, make 1

plot(urban, col = c("grey90", "plum"))
```

## UNESCO Cultual Sites

```{r}
unesco <- read_csv("data/whc-cultural-sites-2021.csv") %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = st_crs("epsg:4326")) %>% 
  filter(states_name_en == "China") %>%
  st_transform(crs = my_crs) %>% 
  dplyr::select(name_en) %>% 
  mutate(ID = row_number()) # define id for joining later

ggplot() + 
  geom_sf(data = china) + 
  geom_sf(data = unesco, 
          color = "blue") 
```



## Define overlaps
### UNESCO + WDPA
```{r}
# vects play better with rasts (both are terra)
unesco_vect <- vect(unesco)

plot(wdpa)
plot(unesco_vect, col = "blue", add = TRUE)
```

```{r}
# use extract to extract what's underneath points
wdpa_overlap <- terra::extract(
  x = wdpa,
  y = unesco_vect) %>% 
  rename(wdpa_overlap = 2) %>% 
  mutate(wdpa_overlap = as.logical(wdpa_overlap)) %>% # change to true/false 
  right_join(unesco, by = "ID") %>%  # join back with unesco data
  dplyr::select(!c(geometry, ID)) %>% # clean df
  relocate(name_en, .before = wdpa_overlap)

kable(wdpa_overlap) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```



### UNESCO + Urban

```{r}
plot(urban, col = c("grey90", "plum"))
plot(unesco_vect, col = "blue", add = TRUE)
```

```{r}
urban_overlap <- terra::extract(
  x = urban,
  y = unesco_vect) %>% 
  rename(urban_overlap = 2) %>% 
  mutate(urban_overlap = as.logical(urban_overlap)) %>%
  right_join(unesco, by = "ID") %>% 
  dplyr::select(!c(geometry, ID)) %>%
  relocate(name_en, .before = urban_overlap)

kable(urban_overlap) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```

## Combine
```{r}
overlaps <- wdpa_overlap %>% 
  left_join(urban_overlap, by = "name_en") 

kable(overlaps) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```

