---
title: "global map"
author: "Nick McManus"
date: '2022-08-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)  # polygons
library(raster) #raster package
library(terra)  # newer/faster/better raster package
library(rnaturalearth)  #administrative boundaries data
library(exactextractr)  # extracting what's underneath polys, points
library(kableExtra)
library(tidyverse)
library(here)
```


## Read in data

This analysis will require data sets of protected areas from the WDPA, urban-classified land use, and sites from the UNESCO World Heritage Convention. WDPA and urban areas are currently excluded from agriculture and energy land use within the current SPARCLE workflow. Because UNESCO WHC sites are protected, these locations must also be excluded, or "locked out" from the SPARCLE process. The purpose of this script is to determine how many cultural heritage sites *do not* overlap with either a WDPA or urban area (at 5km resolution) *and* are larger than 100ha (1 km2) in area. 

### WDPA data

UNEP-WCMC and IUCN (2021), Protected Planet: The World Database on Protected Areas (WDPA) [On-line], Cambridge, UK: UNEP-WCMC and IUCN. Available at: www.protectedplanet.net
```{r}
# read in 5km resolution WDPA layer
wdpa <- rast(here("data", "wdpa_5km_raster.tif"))

# set desired crs to match wdpa layer
my_crs <- st_crs(wdpa)
```

### Urban data

Current (2015) areas classified as urban under Chen et al., 2022
Chen, G., Li, X., & Liu, X. (2022). Global land projection based on plant functional types with a 1-km resolution under socio-climatic scenarios. Scientific Data, 9(1), 125. https://doi.org/10.1038/s41597-022-01208-6
```{r}
# read in 5km resolution urban Chen data
urban <- rast(here("data", "chen_urban_5km.tif")) %>% 
  project(y = wdpa) # make urban and wdpa same crs

```

## UNESCO data

For this dataset, we will only be examining "cultural" heritage sites. The boundaries for sites classified as "natural" or "mixed" are included in the WDPA.

Copyright © 1992 - 2022 UNESCO/World Heritage Centre. All rights reserved.
https://whc.unesco.org/en/syndication
```{r}
# read in cultural heritage sites
unesco <- read_csv(here("data", "whc-sites-2021.csv")) %>% 
  filter(category == "Cultural") %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = st_crs("epsg:4326")) %>% 
  st_transform(crs = my_crs) %>% 
  dplyr::select(name_en, area_hectares, states_name_en) %>% 
  mutate(ID = row_number()) %>%  # define id for joining later
  rename(country = states_name_en)

```

Plot all three together
```{r}
plot(wdpa)
plot(urban, add = TRUE)
plot(unesco, col = "blue", add = TRUE)
```


## Define overlaps

### UNESCO + WDPA

Determine where cultural heritage sites overlap with existing protected areas in the WDPA.
```{r}
# vects play better with rasts (both are terra)
unesco_vect <- vect(unesco)

plot(wdpa)
plot(unesco_vect, col = "blue", add = TRUE)
```

```{r}
# use extract() to extract what's underneath points
wdpa_overlap <- terra::extract(
  x = wdpa,
  y = unesco_vect) %>% 
  # extract what's under the points in the raster
  
  rename(wdpa_overlap = 2) %>%  
  # rename second column to 'urban_overlap'
  
  mutate(wdpa_overlap = as.logical(wdpa_overlap)) %>% 
  # extract() outputs the values of the raster (or a summary thereof)
  # change to true/false 
  
  right_join(unesco, by = "ID") %>% 
  # the df from extract only contains ID numbers and raster values,
  # this joins the dataframe back up with the site names by linking the ID
  # numbers in the two tables
  
  dplyr::select(!c(geometry, ID)) %>% 
  # remove columns we don't care about
  
  relocate(name_en, .before = wdpa_overlap)
  # move columns into desired order
  

kable(wdpa_overlap) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```


### UNESCO + Urban

Determine where cultural heritage sites overlap with existing urban areas. 
```{r}
plot(urban)
plot(unesco_vect, col = "blue", add = TRUE)
```

```{r}
urban_overlap <- terra::extract(x = urban, y = unesco_vect) %>% 
  rename(urban_overlap = 2) %>% 
  mutate(urban_overlap = as.logical(urban_overlap)) %>% 
  right_join(unesco, by = "ID") %>% 
  dplyr::select(!c(geometry, ID)) %>% 
  relocate(name_en, .before = urban_overlap)

# change NAs to FALSE for filtering later
urban_overlap[is.na(urban_overlap)] <- FALSE

kable(urban_overlap) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```

## Combine

We are interested in sites that return FALSE for both urban and WDPA overlaps. Therefore, we must join the previous results. 
```{r}
wdpa_urban_overlap <- wdpa_overlap %>% 
  left_join(urban_overlap, by = NULL) %>%
  relocate(urban_overlap, .before = area_hectares)

# remove NA values for easier filtering later
wdpa_urban_overlap[is.na(wdpa_urban_overlap)] <- FALSE

kable(wdpa_urban_overlap) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```


## Filter for sites of interest

Now, we want to return only cultural heritage sites that do not overlap with either the urban or WDPA layers *and* are at least 100ha in size

```{r}
# filter combined data
unesco_filtered <- wdpa_urban_overlap %>% 
  filter(wdpa_overlap == FALSE,
         urban_overlap == FALSE) %>% 
  filter(area_hectares >= 100)


kable(unesco_filtered) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```

Combine back with original data to get coordinates of filtered sites
```{r}
# read in original UNESCO csv
raw <- read_csv(here("data", "whc-sites-2021.csv")) %>% 
  filter(category == "Cultural")

# join filtered sites with raw csv to assign lat/long data to each unesco site
raw_unesco_joined <- left_join(unesco_filtered, raw, by = c("name_en", "area_hectares"))

# assign CRS and keep only variables of interest
unesco_filtered_geom <- raw_unesco_joined %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = st_crs("epsg:4326")) %>% 
  st_transform(crs = my_crs) %>% 
  dplyr::select(name_en, area_hectares, wdpa_overlap, urban_overlap, country, geometry)

# turn into vector for plotting
unesco_filtered_vect <- vect(unesco_filtered_geom)
```

Plot the results
```{r}
# get country outlines from rnaturalearth
world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  st_transform(crs = crs(wdpa)) # make same crs as rest of data

plot(wdpa)
plot(urban, col = "navyblue", add = TRUE)  #pretty hard to see since so small
plot(world, col = NA, lwd = 0.8, add = TRUE)
plot(unesco_filtered_vect, col = "slateblue4", alpha = 0.8, cex =0.9, add = TRUE)
```


## Analysis

Now we are interested in the size (in hectares) of the remaining cultural heritage sites.

```{r}
# overall stats of remaining cultural sites
mean_size_ha <- mean(unesco_filtered$area_hectares)
median_size_ha <- median(unesco_filtered$area_hectares)
sd_size_ha <- sd(unesco_filtered$area_hectares)
n <- count(unesco_filtered)
min_size <- min(unesco_filtered$area_hectares)
max_size <- max(unesco_filtered$area_hectares)

# how many sites were there originally?
n_raw <- count(raw)
```

Of the `r n_raw` UNESCO cultural heritage sites, `r n` do not overlap with urban or WDPA areas and are at least 100ha large. These remaining sites range in size from `r min_size` to `r max_size`. The mean size is `r mean_size_ha`ha $\pm$ `r sd_size_ha` (mean $\pm$ std dev) with a median size of `r median_size_ha`. Sample statistics summarized by country are returned below:

```{r}
# look by country 
unesco_filtered_stats <- unesco_filtered %>% 
  group_by(country) %>% 
  summarise(mean_area_ha = mean(area_hectares),
            sd_area_ha = sd(area_hectares),
            median_size_ha = median(area_hectares),
            n = n()) %>% 
  relocate(n, .before = mean_area_ha)

# return summary table
kable(unesco_filtered_stats) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```

