---
title: "global map"
author: "Nick McManus"
date: '2022-08-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)  # polygons
library(raster) #raster package
library(terra)  # newer/faster/better raster package
library(rnaturalearth)  #administrative boundaries data
library(exactextractr)  # extracting what's underneath polys, points
library(kableExtra)
library(tidyverse)
```


## Read in data

### WDPA data

Cite info here
```{r}
# read in 5km resolution WDPA layer
wdpa <- rast("data/wdpa_5km_raster.tif")

# set desired crs to match wdpa layer
my_crs <- st_crs(wdpa)
```

### Urban data

Cite info here
```{r}
# read in 5km resolution urban Chen data
urban <- rast("data/chen_urban_5km.tif") %>% 
  project(y = wdpa) # make urban and wdpa same crs

```

## UNESCO data

Cite info here

Only looking at cultural sites for this analysis because natural and mixed have boundaries included in  WDPA layer already
```{r}
# read in cultural heritage sites
unesco <- read_csv("data/whc-sites-2021.csv") %>% 
  filter(category == "Cultural") %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = st_crs("epsg:4326")) %>% 
  st_transform(crs = my_crs) %>% 
  dplyr::select(name_en, area_hectares, states_name_en) %>% 
  mutate(ID = row_number()) %>%  # define id for joining later
  rename(country = states_name_en)

```

Plot all three together
```{r}
plot(wdpa)
plot(urban, add = TRUE)
plot(unesco, col = "blue", add = TRUE)
```


## Define overlaps
### UNESCO + WDPA
```{r}
# vects play better with rasts (both are terra)
unesco_vect <- vect(unesco)

plot(wdpa)
plot(unesco_vect, col = "blue", add = TRUE)
```

```{r}
# use extract() to extract what's underneath points
wdpa_overlap <- terra::extract(
  x = wdpa,
  y = unesco_vect) %>% 
  # extract what's under the points in the raster
  
  rename(wdpa_overlap = 2) %>%  
  # rename second column to 'urban_overlap'
  
  mutate(wdpa_overlap = as.logical(wdpa_overlap)) %>% 
  # extract() outputs the values of the raster (or a summary thereof)
  # change to true/false 
  
  right_join(unesco, by = "ID") %>% 
  # the df from extract only contains ID numbers and raster values,
  # this joins the dataframe back up with the site names by linking the ID
  # numbers in the two tables
  
  dplyr::select(!c(geometry, ID)) %>% 
  # remove columns we don't care about
  
  relocate(name_en, .before = wdpa_overlap)
  # move columns into desired order
  

kable(wdpa_overlap) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```


### UNESCO + Urban
```{r}
plot(urban)
plot(unesco_vect, col = "blue", add = TRUE)
```

```{r}
urban_overlap <- terra::extract(x = urban, y = unesco_vect) %>% 
  rename(urban_overlap = 2) %>% 
  mutate(urban_overlap = as.logical(urban_overlap)) %>% 
  right_join(unesco, by = "ID") %>% 
  dplyr::select(!c(geometry, ID)) %>% 
  relocate(name_en, .before = urban_overlap)

# change NAs to FALSE for filtering later
urban_overlap[is.na(urban_overlap)] <- FALSE

kable(urban_overlap) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```

## Combine

```{r}
wdpa_urban_overlap <- wdpa_overlap %>% 
  left_join(urban_overlap, by = NULL) %>%
  relocate(urban_overlap, .before = area_hectares)

# remove NA values for easier filtering later
wdpa_urban_overlap[is.na(wdpa_urban_overlap)] <- FALSE

kable(wdpa_urban_overlap) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```


## Filter for sites of interest

Now, we want to return only cultural heritage sites that do not overlap with either the urban or WDPA *and* are at least 100ha in size

```{r}
# filter combined data
unesco_filtered <- wdpa_urban_overlap %>% 
  filter(wdpa_overlap == FALSE,
         urban_overlap == FALSE) %>% 
  filter(area_hectares >= 100)


kable(unesco_filtered) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(height = "400px")
```


Combine back with original data to get coordinates of filtered sites
```{r}
# read in raw csv
raw <- read_csv("data/whc-sites-2021.csv") %>% 
  filter(category == "Cultural")

# join filtered unesco sites with raw csv to get lat/long data
raw_unesco_joined <- left_join(unesco_filtered, raw, by = c("name_en", "area_hectares"))

# assign CRS and keep only variables of interest
unesco_filtered_geom <- raw_unesco_joined %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = st_crs("epsg:4326")) %>% 
  st_transform(crs = my_crs) %>% 
  dplyr::select(name_en, area_hectares, wdpa_overlap, urban_overlap, country, geometry)

# turn into vector
unesco_filtered_vect <- vect(unesco_filtered_geom)
```

Plot the results
```{r}
# get country outlines from rnaturalearth
world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  st_transform(crs = crs(wdpa)) # make same crs as rest of data

plot(wdpa)
plot(urban, col = "navyblue", add = TRUE)  #pretty hard to see since so small
plot(world_trans, col = NA, lwd = 0.8, add = TRUE)
plot(unesco_filtered_vect, col = "slateblue4", alpha = 0.8, cex =1.1, add = TRUE)
```



## Analysis

Let's look at the size (in hectares) of the remaining cultural heritage sites

```{r}
# overall stats of remaining cultural sites
mean_size_ha <- mean(unesco_filtered$area_hectares)
median_size_ha <- median(unesco_filtered$area_hectares)
sd_size_ha <- sd(unesco_filtered$area_hectares)
n <- count(unesco_filtered)
mean_size_ha
median_size_ha
sd_size_ha
n


# look by country 
unesco_filtered_stats <- unesco_filtered %>% 
  group_by(country) %>% 
  summarise(mean_area_ha = mean(area_hectares),
            sd_area_ha = sd(area_hectares),
            median_size_ha = median(area_hectares),
            n = n())

```






